Variance reduction for Markov chains via martingale representations

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update

\begin{frontmatter}
\title{Variance reduction for MCMC methods via martingale representations}
\runtitle{Variance reduction for MCMC}
%\thankstext{T1}{This research was }

@@

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation for Markov chains. Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that it's complexity is indeed asymptotically smaller than one of the original MCMC algorithm. The numerical performance of the new method is illustrated  in the case of Gaussian mixtures and Bayesian regression models.
\end{abstract}

@@

Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian statistics. Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$. Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution. Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$. The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand. The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$. Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated. An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance. Methods to achieve this are known as variance
reduction techniques. Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.

@@

One of the popular approaches to variance reduction is the so called control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) + \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.

@@

Unfortunately it is not always possible to generate independent observations distributed according to $\pi$. To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows. It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult. Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case. Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}. At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\). As it was observed in~\cite{henderson1997variance}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\). Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\). For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance. Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.

@@

\paragraph{Notations. }\label{par:notations}
We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution. For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm. For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$. For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
\[
\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,y_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
\]
In the particular case $m=1$ we drop the subscript $y_1$ in that notation. For probability measures $\mu$ and $\nu$ on $\rset^d$ denote by $\tvnorm{\mu-\nu}$ the total variation distance between $\mu$ and $\nu$, that is,
\[
\tvnorm{\mu-\nu}=\sup_{A \in \mathcal{B}(\rset^d)}
|\mu(A)-\nu(A)| \eqsp.
\]
where $\mathcal{B}(\rset^d)$ is a Borel $\sigma$-algebra of $\rset^d$. For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$. Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d}\frac{|f(x)|}{V(x)}$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
\[
d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\limits_{\rset^d}f(x)\,d\mu(x) - \int\limits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
\]

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-\frac{U(x)}{2}}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr),

@@

$\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denotes the density of a $d$-dimensional
standard Gaussian random vector.  The MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form

@@

where \(W\) is a standard \(\mathbb{R}^m\)-valued Brownian motion, \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.

@@

where for all $y \in \rset^d$
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.

>>>>

\begin{frontmatter}
\title{Variance reduction for Markov chains via martingale representations}
\runtitle{Variance reduction for Markov chains}
%\thankstext{T1}{This research was }

@@

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  additive functionals of Markov chains relying on a novel discrete time martingale representation. Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its complexity is indeed asymptotically smaller than one of the naive algorithm. The numerical performance of the new method is illustrated for Markov Chain Monte Carlo (MCMC) methods.
\end{abstract}

@@

Markov chains and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian inference, reinforcement learning, online learning. As an illustration, suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$. Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution. Given an independent identically distributed observations $X_1,\ldots,X_n$ from $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$. The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand. The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$. Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated. An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance. Methods to achieve this are known as variance
reduction techniques. Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for Markov chains, mostly with applications to MCMC algorithms; see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{oates:girolami:chopin:2016}, \cite{south:mira:drovandi:2018}, \cite{brosse2018diffusion} and references therein.

@@

One of the popular approaches to variance reduction is the control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) + \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example \cite[Section~3.2]{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{assaraf1999zero}, \cite{mira2013zero}, and Stein operator in \cite{brosse2018diffusion}.

@@

Unfortunately it is not always possible to generate independent observations distributed according to $\pi$. To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows. It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult. Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case. Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}. At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\). As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\). Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\). For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance. Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.

@@

\paragraph{Notations. }\label{par:notations}
We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution. For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm. For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$. For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
\[
\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,y_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
\]
In the particular case $m=1$ we drop the subscript $y_1$ in that notation. For probability measures $\mu$ and $\nu$ on $\rset^d$ denote by $\tvnorm{\mu-\nu}$ the total variation distance between $\mu$ and $\nu$, that is,
\[
\tvnorm{\mu-\nu}=\sup_{A \in \mathcal{B}(\rset^d)}
|\mu(A)-\nu(A)| \eqsp.
\]
where $\mathcal{B}(\rset^d)$ is a Borel $\sigma$-algebra of $\rset^d$. For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$. Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} |f(x)|/V(x)$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
\[
d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\nolimits_{\rset^d}f(x)\,d\mu(x) - \int\nolimits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
\]

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see \cite[Theorem~1.3.6]{moulines2018})
and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= Z^{-1} \rme^{-U(x)/2}, \quad Z= \int_{\mathbb{R}^{d}} \rme^{-U(x)/2}\, \rmd x,  
\end{equation}
provided that \(Z < \infty\). 
This method is usually referred to as Unadjusted Langevin Algorithm (ULA).
Denote by \(W\) the standard \(\mathbb{R}^m\)-valued Brownian motion.
The Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr) \,.

@@

The MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form

@@

where \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.

@@

where for all $y \in \rset^d$, 
$\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right]$, $r, k\in \mathbb{N}$.
